Towards practical dynamic induction control of wind farms: analysis of optimally controlled wind-farm boundary layers and sinusoidal induction control of first-row turbines

### 2018-02-06 draft.tex
Comments JM: section 3 (minus 3.5)

The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in MM17, and performs numerical experiments to uncover some of the characteristics of these control signals. The aim of this analysis is to increase our overall understanding of how power is increased in optimally controlled wind-farm boundary layers, i.e. how the potential for power gains depends on the location of a turbine within the farm, whether optimal controls are fully tuned to specific flow events or contain flow-invariant features, and whether the control signals in neighboring turbines are correlated or can be used independently. This increased understanding will be used as a stepping stone towards the design of practical control strategies that do not require the computationally prohibitive LES-based methodology of MM17.

First, the thrust coefficient signals themselves are analyzed in Section~\ref{sec:analysis_thrust}. Second, the optimized thrust coefficients are applied only to subsets of turbine rows in Section~\ref{sec:application_subset}. In this way, the interdependency of optimized thrust coefficients in different rows can be evaluated. Third, additional optimal control simulations, in which only one single active row is optimized, are discussed in Section~\ref{sec:optimization_single}. These optimizations provide an indication on how increased power potential is distributed among the rows, and allows to compare the resulting single-row optimized thrust coefficients with the fully cooperative coefficients from case MM17. Fourth, Section~\ref{sec:modified_thrust} evaluates the dependency of optimized thrust coefficients on specific flow conditions. This is done by running test simulations in which the coefficients are modified to eliminate any possible correlations to specific flow events at given times. Finally, Section~\ref{sec:analysis_discussion} discusses the main conclusions from the abovementioned sections, and summarizes the lessons learned.

@@

Figure~\ref{fig:controls} illustrates the time evolution of some of  the thrust coefficients $\cthat$ in the C3t5 case. The figure illustrates that, for all rows but the last one (i.e. R12), $\cthat$ varies significantly in time, and that the amplitudes and frequencies of these variations are somewhat higher in the upstream rows of the farm. In contrast, row 12 features only minor unsteadiness at lower frequencies, and has an increased mean value of $\cthat$. This relatively steady behavior of the last row can be explained by the fact that there are no further downstream turbines that could benefit from row 12 actively influencing local flow conditions, hence the row optimizes its own power greedily. The increase in mean $\cthat$ can be explained based on Figure~\ref{fig:ct_sweep}, which illustrates the power extraction as a function of steady $\cthat$ for unwaked turbines, subject to identical turbulent inlet as in case C3t5. Although momentum theory predicts maximal power extraction for steady uniform inflow at $\cthat = 2$, the actual optimal steady value for the ADM at the current spatial resolution lies somewhat higher at $\cthat \approx 2.4$, for which power extraction is about 1.4\% higher than at $\cthat = 2$. This behavior is related to the overprediction in ADM power due to the diffuse turbine representation on typical simulation grids: the mass flow through the rotor disk at $\cthat > 2$ is slightly too high compared to momentum theory, resulting in a  shift of optimal $\cthat$ towards somewhat higher values. Although the linear fit $C_P' = a \cthat$ introduced in \cite{munters2017optimal}, Appendix~A eliminates the error in maximal power extraction, it does not correct the value of the optimal $\cthat$ (note that this could be achieved through a more complex relation between $C_P'$ and $\cthat$). Even though this explains the slight increase in mean $\cthat$ for greedily optimized last-row turbines, this does not undermine the overall power gains of the overinductive optimal control cases in MM17, which all feature increases well over 10\%. Returning to the more complex thrust coefficients in the other rows it is worth noting that, based on the current dataset, no statistically significant correlations between thrust coefficients of different turbines could be found. Furthermore, attempts towards linking thrust coefficient dynamics to upstream flow measurements (e.g. velocities, shear or kinetic energy) through linear regression models and random forest regressors have been unsuccessful to date.

@@

Figure~\ref{fig:spectra_thrust_cthat}a,b illustrate power spectral densities of the row-averaged thrust forces and thrust coefficients respectively. The figure illustrates that the variances of both the thrust coefficients and their resulting forces are highest in row 1. Further downstream, rows 2 to 11 have very similar spectral behavior, and row 12 shows significantly lower variability. The high-frequency slopes of around $-5$ observed both for $f_T$ and $\cthat$ indicate that force variability on short timescales is caused mainly by thrust coefficient variations, whereas the slower thrust force dynamics tend more to a $-5/3$ slope, suggesting that these are governed by the unsteadiness in the turbulence instead. Note that, even though the spectra for all rows except row 12 collapse at frequencies below $0.05$ Hz, the first-row spectrum shows a small peak at $f \approx 0.02 \dots 0.03$ Hz ($f D / U_\infty \approx 0.2 \dots 0.3$) as indicated by the purple arrow, which also translates to an increased spectral content of the first row thrust force at these frequencies. It will be shown later in this paper that variations in the thrust coefficient around this frequency are directly related to increased power extraction.

@@

In order to further study how the optimal controls increase overall wind-farm power, Figure~\ref{fig:multirow} illustrates power extraction resulting from applying a subset of the optimal controls to specific turbine rows only. Figure~\ref{fig:multirow}a depicts simulation results for which the  optimized controls are applied \emph{only to one specific row}, with the thrust coefficient in all other rows kept at the reference value of $\cthat = 2$. From the figure it can be seen that only for the controls of the first row (R1) this results in a significant power increase in rows 2 and 3. This indicates that the optimal controls, as generated by the optimization at the wind-farm level, depend on the precise flow conditions caused by upstream control actions, and can hence only be applied independently for the first row, which has no upstream dependence by definition.

Figure~\ref{fig:multirow}b shows results from simulations in which the controls are applied \emph{for all rows up to a certain row}, i.e. R1--R3 indicates the application of optimized controls generated by case C3t5 to rows 1, 2, and 3. An interesting observation from this figure is that , for any row $i$ except the last one, the power potential as observed in case C3t5 is almost fully recovered by only applying the optimal controls up to row $i-1$. This suggests that self-optimization is very limited: the optimal controls for a given turbine are designed to create favorable flow conditions in the downstream rows instead of increasing local power. Furthermore, although the discussion in the previous paragraph has shown that downstream controls are optimized with the upstream actions in mind, the converse is not true: upstream control actions do not require a specific downstream response in order to increase power in downstream rows.

@@

\subsection{Optimization of single active turbine rows}\label{sec:optimization_single}
The previous section has shown that, based on the full-farm optimization case, the first-row controls can be applied independently, whereas this does not work for the downstream rows. To further quantify the potential for increasing wind-farm power in each row of turbines, the current section considers a set of additional optimal control cases in which \emph{only a single active row is optimized}, with all other rows remaining passive. Furthermore, by comparing the optimized controls of these cases with the full-farm optimization case C3t5, the degree of cooperation between turbines can be assessed. Note that the current single-row optimal control is not equivalent to greedy control: the optimizer still aims to increase aggregate farm power by taking into account wake interactions with downstream turbines. Furthermore, in contrast to the single-row control simulations from the previous section (i.e. in Figure~\ref{fig:multirow}a), the current optimizations will yield controls that are \emph{explicitly designed} to increase power given that all other rows are passive. To limit computational costs, the additional optimizations are only performed for a single time window.

Figure~\ref{fig:single_row_opt} illustrates the relative increase in wind-farm power extraction for each of the twelve individually optimized control cases. The optimization is run until the continuous adjoint gradient accuracy prevents further progress in the vicinity of a local minimum. \red{Upon interpreting the actual values from the figure, it is important to note that the reported power gain covers the full optimization horizon $T$, and is hence affected by finite-horizon effects. Furthermore, the first window of an optimal control simulation, as considered here, contains an initial dead zone corresponding to the wake advection lag before upstream turbines start influencing their downstream neighbors. This tends to reduce gains compared to later time windows. Nevertheless, the relative order of the different cases still provides information that can be generalized to full optimal control studies with multiple windows and longer time horizons. (WM: dit eventueel weglaten?)}

The figure shows that the first row (R1) holds by far the most promise for optimizing wind-farm power. This is not surprising as R1 produces the most power of all rows, and typically leaves the deepest wakes, causing second-row turbines to perform  poorly in aligned wind-farm layouts (see, e.g., \citealp{porte2013numerical,nilsson2015large}; and \citealp{stevens2016effects}). At the other end of the spectrum, the last row (R12) is the least fruitful. The intermediate rows (R2--R11) lie closer together, with the general trend being that potential is somewhat decreased with downstream distance into the wind farm, although this decrease is not monotonous.

@@

Figure~\ref{fig:gains_singlerow} illustrates the row-wise relative power increase matrix for each of the single-row optimization cases. The figure indicates that, for each of the optimization cases, the largest power increase is observed in the first row downstream of the active turbine (i.e. R$_{\rm i+1}$), and that the influence on row R$_{\rm i+3}$ is limited. This is explained by the fact that the finite optimization horizon used in MM17 (i.e. $T = 240$ s) allows for more interactions with directly neighboring turbines than with those located further downstream. Furthermore, except for the optimal control case of the last row (R12), self-optimization is virtually non-existent: power gains are achieved by modifying the flow to yield more favorable conditions in the downstream rows.

@@

Turning to the thrust dynamics, Figure~\ref{fig:control_C3t5_singlerow} shows the optimized thrust coefficients for the entire optimization horizon $T$ in several turbine rows for the single-row (gray) and full-farm (black) optimizations. Interestingly, the first-row thrust coefficient is virtually identical, regardless of whether downstream turbines are active or not. Although similarities can also be observed for the downstream rows, the correspondence is much weaker than in the first row. This can be explained by the fact that, in contrast to the first row, downstream turbines experience different local flow conditions based on whether the upstream turbines are active or not.

\begin{figure}
	\centering
	\includegraphics[width=0.8\textwidth]{fullfarm_singlerow.eps}
	\caption{Optimized thrust coefficients in several rows for case C3t5 (black) and single-row optimal control cases (gray) for the full optimization horizon $T = 240$ s.\label{fig:control_C3t5_singlerow}}
\end{figure}

The current observations further confirm the conclusions from previous section: optimal control actions depend to some degree on the local flow conditions. Furthermore, previous section showed that first-row controls still lead to power gains in the absence of downstream control response. Here, as indicated by the similarity in first-row controls for single-row and full-farm optimization cases, it is shown that, for the first row, the optimizer is actually indifferent to whether or not downstream rows are actively controlled.

\subsection{Modification of thrust coefficient signals}\label{sec:modified_thrust}
The observations from previous sections illustrate that, at least to some degree, the optimized thrust coefficients are tuned to local flow conditions. In the current section, the possibility of whether the coefficients contain traits that are independent of flow conditions is nonetheless investigated. To this end, the optimized thrust coefficients are modified in such a way that correlations between them and specific flow events are eliminated. This is done in two independent test cases.

@@

The figure shows similar behavior for each of the modified control cases: only in the second row (R2), a consistent increase in power extraction can be observed. Combining this observation with the findings of previous sections, i.e. that power gains in a given row are caused by upstream control action instead of self-optimization, the current test cases suggest the presence of flow-invariant	features in the control signals of the first row. Note however that the full power gain in the second row is only partially attained, indicating that also the first turbine row reacts to the specific flow conditions. The small power increase in the last row (R12) can again be explained by the slightly higher mean value in $\cthat$. Based on the current data, no consistent gains in any of the other rows are found.

>>>>

The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in MM17, and performs numerical experiments to uncover some of the characteristics of these control signals.

First, the thrust coefficient signals themselves are analyzed in Section~\ref{sec:analysis_thrust}. Second, the optimized thrust coefficients are applied only to subsets of turbine rows in Section~\ref{sec:application_subset}. In this way, the interdependency of optimized thrust coefficients in different rows can be evaluated. Third, additional optimal control simulations, in which only one single active row is optimized, are discussed in Section~\ref{sec:optimization_single}. These optimizations provide an indication on how increased power potential is distributed among the rows, and allows to compare the resulting single-row optimized thrust coefficients with the fully cooperative coefficients from case MM17. Fourth, Section~\ref{sec:modified_thrust} evaluates the dependency of optimized thrust coefficients on the actual turbulent flow realization. Finally, Section~\ref{sec:analysis_discussion} discusses the main conclusions from the abovementioned sections, and summarizes the lessons learned.

@@

Figure~\ref{fig:controls} illustrates the time evolution of some of  the thrust coefficients $\cthat$ in the C3t5 case. The figure shows that, for all rows but the last one (i.e. R12), $\cthat$ varies significantly in time, and that the amplitudes and frequencies of these variations are somewhat higher in the upstream rows of the farm. In contrast, row 12 features only minor unsteadiness at lower frequencies, and has an increased mean value of $\cthat$. This relatively steady behavior of the last row can be explained by the fact that there are no further downstream turbines that can benefit from row 12 actively influencing local flow conditions, hence the row optimizes its own power only. The increase in mean $\cthat$ in row 12 can be explained based on Figure~\ref{fig:ct_sweep}, which shows the power extraction as a function of steady $\cthat$ for unwaked turbines, subject to identical turbulent inlet as in case C3t5. Although momentum theory predicts maximal power extraction for steady uniform inflow at $\cthat = 2$, the actual optimal steady value for the ADM at the current spatial resolution lies somewhat higher at $\cthat \approx 2.4$, for which power extraction is about 1.4\% higher than at $\cthat = 2$. This behavior is related to the overprediction in ADM power due to the diffuse turbine representation on typical simulation grids: the mass flow through the rotor disk at $\cthat > 2$ is slightly too high compared to momentum theory, resulting in a  shift of optimal $\cthat$ towards somewhat higher values. Although the linear fit $C_P' = a \cthat$ introduced in \cite{munters2017optimal}, Appendix~A eliminates the error in maximal power extraction, it does not correct the value of the optimal $\cthat$ (note that this could be achieved through a more complex relation between $C_P'$ and $\cthat$). Returning to the more complex thrust coefficients in the other rows it is worth noting that, based on the current dataset, no statistically significant correlations between thrust coefficients of different turbines could be found. Furthermore, attempts towards linking thrust coefficient dynamics to upstream flow measurements (e.g. velocities, shear or kinetic energy) through linear regression models and random forest regressors have been unsuccessful to date.

@@

Figure~\ref{fig:spectra_thrust_cthat}a,b shows row-averaged power spectral densities of the thrust forces and thrust coefficients respectively. The figure shows that the variances of both the thrust coefficients and their resulting forces are highest in row 1. Further downstream, rows 2 to 11 have very similar spectral behavior, and row 12 shows significantly lower variability. The high-frequency slopes of around $-5$ observed both for $f_T$ and $\cthat$ indicate that force variability on short timescales is caused mainly by thrust coefficient variations, whereas the slower thrust force dynamics tend more to a $-5/3$ slope, suggesting that these are governed by the unsteadiness in the turbulence instead. Note that, even though the spectra for all rows except row 12 collapse at frequencies below $0.05$ Hz, the first-row spectrum shows a small peak at $f \approx 0.02 \dots 0.03$ Hz ($f D / U_\infty \approx 0.2 \dots 0.3$) as indicated by the purple arrow. It will be shown later in this paper that variations in the thrust coefficient around this frequency are directly related to increased power extraction.

@@

In order to further study how the optimal controls increase overall wind-farm power, Figure~\ref{fig:multirow} shows power extraction resulting from applying a subset of the optimal controls to specific turbine rows only. Figure~\ref{fig:multirow}a depicts simulation results for which the  optimized controls are applied \emph{only to one specific row}, with the thrust coefficient in all other rows kept at the reference value of $\cthat = 2$. From the figure it can be seen that only for the controls of the first row (R1) this results in a significant power increase in rows 2 and 3. This indicates that the optimal controls, as generated by the optimization at the wind-farm level, react to the precise flow conditions caused by upstream control actions, and can hence only be applied independently for the first row, which has no upstream dependence by definition.

Figure~\ref{fig:multirow}b shows results from simulations in which the controls are applied \emph{for all rows up to a certain row}, i.e. R1--R3 indicates the application of optimized controls generated by case C3t5 to rows 1, 2, and 3. An interesting observation from this figure is that , for any row $i$ except the last one, the power potential as observed in case C3t5 is almost fully recovered by only applying the optimal controls up to row $i-1$. This suggests that self-optimization is very limited: the optimal controls for a given turbine are designed to create favorable flow conditions in the downstream rows instead of increasing local power. Furthermore, although the discussion in the previous paragraph has shown that downstream controls are optimized with the upstream actions in mind, the converse is not true: upstream control actions do not require a specific downstream response in order to increase power in that downstream row.

@@

\subsection{Optimization of single active turbine rows}\label{sec:optimization_single}
The previous section has shown that, based on the full-farm optimization case, the first-row controls can be applied independently from other turbine controls, whereas this does not work for the downstream rows. To further quantify the potential for increasing wind-farm power in each row of turbines, the current section considers a set of additional optimal control cases in which \emph{only a single active row is optimized}, with all other rows remaining passive. Furthermore, by comparing the optimized controls of these cases with the full-farm optimization case C3t5, the degree of cooperation between turbines can be assessed. Note that the current single-row optimal control is not equivalent to greedy control: the optimizer still aims to increase aggregate farm power by taking into account wake interactions with downstream turbines. Furthermore, in contrast to the single-row control simulations from the previous section (i.e. in Figure~\ref{fig:multirow}a), the current optimizations will yield controls that are \emph{explicitly designed} to increase power given that all other rows are passive. To limit computational costs, the additional optimizations are only performed for a single time window.

Figure~\ref{fig:single_row_opt} shows the relative increase in wind-farm power extraction for each of the twelve individually optimized control cases. The optimization is run until the continuous adjoint gradient accuracy prevents further progress in the optimization. Upon interpreting the actual values from the figure, it is important to note that the reported power gain covers the full optimization horizon $T$, and is hence affected by finite-horizon effects. Furthermore, the first window of an optimal control simulation, as considered here, contains an initial dead zone corresponding to the wake advection lag before upstream turbines start influencing their downstream neighbors. This tends to reduce gains compared to later time windows. Nevertheless, the relative order of the different cases still provides information that can be generalized to full optimal control studies with multiple windows and longer time horizons.

The figure shows that the first row (R1) holds by far the most promise for optimizing wind-farm power. This is not surprising as R1 produces the most power of all rows, and typically leaves the deepest wakes, causing second-row turbines to perform  poorly in aligned wind-farm layouts (see, e.g., \citealp{porte2013numerical,nilsson2015large}; and \citealp{stevens2016effects}). At the other end of the spectrum, the last row (R12) is the least useful. The intermediate rows (R2--R11) lie closer together, with the general trend being that the potential is somewhat decreased with downstream distance into the wind farm, although this decrease is not monotonous.

@@

Figure~\ref{fig:gains_singlerow} illustrates the row-wise relative power increase matrix for each of the single-row optimization cases. The figure indicates that, for each of the optimization cases, the largest power increase is observed in the first row downstream of the active turbine (i.e. R$_{\rm i+1}$), and that the influence on row R$_{\rm i+3}$ is limited. This is explained by the fact that the finite optimization horizon used in MM17 (i.e. $T = 240$ s) allows for more interactions with directly neighboring turbines than with those located further downstream. Furthermore, except for the optimal control case of the last row (R12), self-optimization is virtually non-existent: power gains are achieved by modifying the flow to yield more favorable conditions for downstream rows.

@@

\subsection{Modification of thrust coefficient signals}\label{sec:modified_thrust}
The observations from previous sections illustrate that, at least to some degree, the optimized thrust coefficients are tuned to local flow conditions. In the current section, the possibility of whether the coefficients contain traits that are independent of flow conditions is investigated. To this end, the optimized thrust coefficients are modified in such a way that correlations between them and specific flow events are eliminated. This is done in two independent test cases.

@@

The figure shows similar behavior for each of the modified control cases: only in the second row (R2), a consistent (though small) increase in power extraction can be observed. This suggests the presence of flow-invariant	features in the control signals of the first row. Note however that the full power gain in the second row is only partially attained, indicating that also the first turbine row reacts to the specific flow conditions.
